NURM: a tool to locate numerical differences in pipelines

### 2019-09-25 paper.tex
Comments on Experiments and results

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{amsmath}
\DeclareMathOperator*{\argmin}{argmin}

%%%% Commands %%%%
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note: #1\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{blue}\textbf{From Tristan:}#1\color{black}}

@@

Indeed, the propagation and amplification of these tiny perturbations by analysis 
may cause a significant reproducibility issue. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

@@

We used our tool to evaluate the effect of the operating system on results 
produced by the pipelines from the Human Connectome Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is a significant effort that will acquire and analyze 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open source license.

@@

Also, the main goals of HCP FreeSurfer pipeline are:
(1) improve the robustness of brain extraction,
(2) fine tune image registration,
(3) accurately place the surface images with high resolution data,
(4) perform FreeSurfer's standard folding-based surface registration 
to their surface atlas.

@@

HCP data consists of brain images collected from an individual healthy person, 
in the age range between 22-35 years~\cite{van2013wu}. Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
Such a directory containing the HCP data is denoted by the term ``subject".
In this paper, we randomly selected 20 unprocessed subjects from 
the HCP data release S500. All the subjects were scanned using the 3-tesla 
type of MRI scanners.

@@

The data were processed in the order PreFreeSurfer and FreeSurfer. 
Unprocessed subject data used as the input data for PreFreeSurfer and its output 
results used as the input data for FreeSurfer analysis in each operating system separately.
Through all the subjects, first, two runs were made on the same operating systems 
to be sure that identified differences are not coming from the operations with randomness.
Then we identified the processes in the pipelines that are responsible for 
the creation of differences using the NURM-tool.

It should be noted that some files may have the same data but 
different metadata in headers, including creation time, modification time and date, 
which turns to non-identical checksums. 
So, we added some other functions for the file comparison to be sure that differences 
are in part of the file data. 
Depending on the file type, we used functions from FreeSurfer software package to determine 
whether two files differ, including  
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} for comparing volumes, surface-based files, 
and lta files (transforms) respectively. 
These functions check data based on the volume dimentions, 
resolutions, pixel intensities,  
the number of surface vertices and neighbors, and geometry information. 
It then returns files differ if there be differences in any of these parameters.
Also, text files are compared using a custom function that compares the content line by line.

@@

Two types of differences can occur in the subjects due to the 
differences in the operating systems. One is inter-OS differences caused by the 
operating system library updates and the other type, intra-OS differences 
occur as a result of the pseudo-random processes used in the 
pipelines.

In our experiments, we identified the processes that create differences due to 
intra-OS differences.  
For example, process \emph{mris\_curvature}
computes mean and Gaussian curvatures of a cortical surface using 
the random variable in FreeSurfer.
To avoid such randomness we 
initialized random number generators using a seed state in 
the command line in which pipeline script calls the program.
This ensures that results are the same from run to run.
Therefore, NURM-tool can identify the processes in the pipelines 
that create inter-OS differences.

\subsubsection{Inter-OS binary differences}

In a previous study~\cite{Scaria2017}, we showed that the HCP pre-processing 
pipelines~\cite{glasser2013} were 
sensitive to operating system variation.

Figure~\ref{fig:fnirt_result} shows differences in the files 
that are created as a part of ACPC-Alignment in PreFreeSurfer.
The square pixels in the brain in the checkerboard images show the differences between results in two OSes. 
The RMSE (root mean squared error) value of results is 0.029 and the dice coefficient is 0.125.
In the next subsection, it is shown that such misalignment is created due to 
non-linear registration in \emph{FNIRT} process in PreFreeSurfer pipeline.

Furthermore, Figure 
\ref{fig:tissue_class} illustrates the heatmap binarized differences 
of the brain tissue segmentation across OSes in FreeSurfer pipeline.
These differences are accumulated from 20 subject results.
So, the magnitude of differences is in the range of 1-20 as the total number of subjects.
The darker blue pixels show differences among less subjests and 
the brighter blue pixels are related to the differences among more subjects.

>>>>

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{amsmath}
\usepackage{ulem}
\DeclareMathOperator*{\argmin}{argmin}

%%%% Commands %%%%
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note: #1\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{blue}\textbf{From Tristan:}#1\color{black}}
\newcommand{\english}[1]{\uwave{#1}}

@@

Indeed, the propagation and amplification of these tiny perturbations by analysis 
may cause a significant reproducibility issue. \tristan{All the above is introduction and should be moved there.
The introduction of this experiment section should explain:
\begin{enumerate}
  \item What is the goal of the experiment
  \item What are the expected results 
  \item Any required context that is specific to the experiment and not described before
  \item The main components of the section: ``This section describes 
  the datasets, pipelines, etc used in the experiments."
\end{enumerate}
}
We focus on the differences 
created by different versions of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

@@

We used our tool to evaluate the effect of the operating system on results 
produced by the pipelines from the Human Connectome Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is a significant effort that will acquire \tristan{check future tense} and analyze 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open source license.

@@

Also, the main goals of HCP FreeSurfer pipeline are:
(1) improve the robustness of brain extraction,
(2) fine tune image registration,
(3) accurately place the surface images with high resolution data,
(4) perform FreeSurfer's standard folding-based surface registration 
to their surface atlas.
\tristan{The two previous paragraphs should focus on a technical description of the pipelines, 
possibly with a workflow figure, 
and refer to ref [12] for any image processing detail.}

@@

HCP data consists of brain images collected from an individual healthy human subjects, 
in the age range between 22-35 years~\cite{van2013wu}. Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
Such a directory containing the HCP data is denoted by the term ``subject".
In this paper, we randomly selected 20 unprocessed subjects from 
the HCP data release S500. All the subjects were scanned using the 3-tesla 
type of MRI scanners.
\tristan{remove details about age range and directory. Instead, explain the type of
imaging that is available: T1 (multiple instances), DWI, functional.}

@@

The data were processed with PreFreeSurfer and FreeSurfer. 
\english{Unprocessed subject data used as the input data for PreFreeSurfer and its output 
results used as the input data for FreeSurfer analysis in each operating system separately}.
Each subject was processed twice on the same operating system to detect
variability coming from pseudo-random operations. 
Then we used NURM-tool to identify the processes in the pipelines that
created differences.

\tristan{This paragraph should be merged with the sentence on pseudo-randomness. 
Prose should also be simplified and made more direct.}
Within-OS variability may also be coming from other factors than pseudo-randomness. For instance, 
metadata may differ due to the presence of timestamps, file paths and other 
execution-specific information. \tristan{in the tool description, talk about the possibility to add distance functions 
that are specific to file types and are computed in addition to checksums.}
To address this issue, we created distance functions \english{for the file comparison to be sure that differences 
are in part of the file data}. 
\tristan{summarize the following sentences, keep it high level, and provide a link to the
metrics file that you used.}
Depending on the file type, we used functions from FreeSurfer software package to determine 
whether two files differ, including  
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} for comparing volumes, surface-based files, 
and lta files (transforms) respectively. 
These functions check data based on the volume dimentions, 
resolutions, pixel intensities,  
the number of surface vertices and neighbors, and geometry information. 
It then returns files differ if there be differences in any of these parameters.
Also, text files are compared using a custom function that compares the content line by line.

@@

Two types of differences can occur in the subjects due to the differences
in the operating systems. One is inter-OS differences caused by the
operating system library updates and the other type, intra-OS differences
occur as a result of the pseudo-random processes used in the pipelines.
\tristan{Move this up to intro and/or experiments, replace "inter-" by
"between-" and "intra-" by "within-"}

\subsubsection{Within-OS binary differences}

In our experiments, we identified the processes that create differences due to 
intra-OS differences.  
For example \tristan{list the processes exhaustively}, process \emph{mris\_curvature}
computes mean and Gaussian curvatures of a cortical surface using 
the random variable in FreeSurfer.
To avoid such randomness we 
initialized random number generators using a seed state in 
the command line in which pipeline script calls the program.
This ensures that results are the same from run to run.
Therefore, NURM-tool can identify the processes in the pipelines 
that create inter-OS differences. \tristan{sharpen a bit the writing}

\subsubsection{Between-OS binary differences}

\tristan{Before showing a particular case, you should describe 
an overview of the differences:
\begin{enumerate}
  \item What is the frequency of binary differences and where do they happen in the pipelines? Do they happen for all subjects?
  \item What is the magnitude of the differences? Explain the metrics, report them across subjects for different files.
  \item Give visual examples.
\end{enumerate}

In a previous study~\cite{Scaria2017}, we showed that the HCP pre-processing 
pipelines~\cite{glasser2013} were 
sensitive to operating system variation \tristan{this is intro}.

Figure~\ref{fig:fnirt_result} shows differences in the files 
that are created as a part of ACPC-Alignment in PreFreeSurfer. \tristan{add link to gif animation}
The square pixels in the brain in the checkerboard images show the differences between results in two OSes. 
The RMSE (root mean squared error) value of results is 0.029 and the dice coefficient is 0.125.

Furthermore, Figure 
\ref{fig:tissue_class} illustrates the sum of binarized differences 
of the brain tissue segmentation across OSes in FreeSurfer pipeline, for the 20 subjects \tristan{add a color 
scale to the figure}.
These differences are accumulated from 20 subject results.
So, the magnitude of differences is in the range of 1-20 as the total number of subjects.
The darker blue pixels show differences among less subjests and 
the brighter blue pixels are related to the differences among more subjects.
